% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{agujournal2019}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother


\usepackage{longtable,booktabs,array}
\newcounter{none} % for unnumbered tables
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother


% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}



\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}



 


\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Machine Learning in Materials Processing \& Characterization},
  pdfauthor={Philipp Pelz},
  pdfkeywords={Machine Learning, Materials Science, Materials
Processing, Materials Characterization, Deep Learning, Microstructure
Analysis, Process Optimization},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}



\draftfalse

\begin{document}
\title{Machine Learning in Materials Processing \& Characterization}

\authors{Philipp Pelz\affil{1}}
\affiliation{1}{Materials Science and Engineering, }
\correspondingauthor{Philipp Pelz}{}


\begin{abstract}
This course teaches how machine learning can be applied to experimental
data from materials processing and characterization. The focus lies on
images, spectra, time-series, and processing parameters, and on
understanding how physical data formation interacts with learning
algorithms. Students learn to build robust, uncertainty-aware ML
pipelines for real experimental workflows, avoiding common pitfalls such
as data leakage, overfitting, and spurious correlations.
\end{abstract}

\section*{Plain Language Summary}
Students learn how to use machine learning on real materials data such
as microscopy images, spectra, and processing logs. The course
emphasizes understanding the physics behind the data, choosing suitable
ML models, and assessing uncertainty and reliability rather than just
maximizing accuracy.




\section{Machine Learning in Materials Processing \&
Characterization}\label{machine-learning-in-materials-processing-characterization}

\textbf{5th Semester -- 5 ECTS · 2h lecture + 2h exercises per week}\\
\emph{Coordinated with ``Mathematical Foundations of AI \& ML'' (MFML)\\
and ``Materials Genomics'' (MG)}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{Synergy Map}\label{synergy-map}

\begin{itemize}
\tightlist
\item
  \textbf{MFML} provides the mathematical spine: loss functions, neural
  networks, generalization, uncertainty, Gaussian Processes.
\item
  \textbf{This course (ML-PC)} applies these concepts to
  \emph{experimental} data: images, spectra, and processing signals.
\item
  \textbf{Materials Genomics} focuses on crystal structures, databases,
  and discovery.
\end{itemize}

ML-PC is therefore \textbf{application-driven}, not algorithm-driven.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{Companion books}\label{companion-books}

\begin{itemize}
\tightlist
\item
  Sandfeld (2024): Materials Data Science
\end{itemize}

\subsection{Week-by-Week Curriculum (14
weeks)}\label{week-by-week-curriculum-14-weeks}

\subsubsection{Unit I --- Experimental Data as a Learning Problem (Weeks
1--3)}\label{unit-i-experimental-data-as-a-learning-problem-weeks-13}

\paragraph{Week 1 -- What makes materials data
special?}\label{week-1-what-makes-materials-data-special}

\emph{Lecture: Tuesday, 14.04.2026, 14:15-15:45 \textbar{} Exercise:
Thursday, 16.04.2026, 16:15-17:45}

\textbf{Slides:}
\href{https://pelzlab.science/public_presentations/ml_for_characterization_and_processing/unit01_intro/01_intro.html}{Open}

\begin{itemize}
\tightlist
\item
  Types of experimental data: micrographs, EBSD, EDS, EELS, XRD, process
  logs, thermal histories.
\item
  PSPP (Processing--Structure--Property--Performance) as a \emph{data
  dependency graph}.
\item
  Why ML failure modes are common in experimental science.
\end{itemize}

\textbf{Summary:} This unit introduces the transition from classical
physics-based modeling to data-driven discovery in materials science. We
explore the unique challenges of experimental materials data, including
its multi-modal nature, high acquisition cost, and the fundamental
Processing-Structure-Property-Performance (PSPP) relationships. Key
concepts include data scales, measurement uncertainty, and the CRISP-DM
process adapted for scientific workflows.

\textbf{Exercise:}\\
Inspect real microscopy and process datasets; identify sources of bias
and noise.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\paragraph{Week 2 -- Physics of data
formation}\label{week-2-physics-of-data-formation}

\emph{Lecture: Tuesday, 21.04.2026, 14:15-15:45 \textbar{} Exercise:
Thursday, 23.04.2026, 16:15-17:45}

\textbf{Slides:}
\href{https://pelzlab.science/public_presentations/ml_for_characterization_and_processing/unit02_physics_of_data/01_intro.html}{Open}

\begin{itemize}
\tightlist
\item
  Image and signal formation in characterization: resolution, contrast,
  artifacts.
\item
  Sampling, aliasing, noise as \emph{physical priors} (not preprocessing
  tricks).
\item
  Relation to MFML refresher on PCA and covariance.
\end{itemize}

\textbf{Summary:} This unit bridges the gap between the physical process
of data acquisition and the mathematical tools used to describe it. We
analyze how signals are formed in characterization tools and how
physical constraints (resolution, noise, sampling) act as priors for
learning. We then introduce Principal Component Analysis (PCA) and
Singular Value Decomposition (SVD) as fundamental techniques for
discovering low-dimensional structure in high-dimensional experimental
datasets.

\textbf{Exercise:}\\
Fourier inspection of micrographs; effects of sampling and filtering.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\paragraph{Week 3 -- Data quality, labels, and
leakage}\label{week-3-data-quality-labels-and-leakage}

\emph{Lecture: Tuesday, 28.04.2026, 14:15-15:45 \textbar{} Exercise:
Thursday, 30.04.2026, 16:15-17:45}

\textbf{Slides:}
\href{https://pelzlab.science/public_presentations/ml_for_characterization_and_processing/unit03_data_quality/01_intro.html}{Open}

\begin{itemize}
\tightlist
\item
  Annotation uncertainty and inter-annotator variance.
\item
  Train/test leakage in materials workflows.
\item
  Why ``good accuracy'' often means a broken pipeline.
\end{itemize}

\textbf{Summary:} This unit covers the often-overlooked half of an ML
pipeline: data integrity, validation, and how performance is measured.
We start with the measurement chain and systematic \textbf{data
cleaning} --- handling missing values, outliers, and duplicates with a
``fix at source'' mindset. We then build the \textbf{transformation
toolbox}: centering, min--max and z-score scaling, physics-aware
non-dimensionalisation, log transforms, differentiation, and
frequency-domain views (FFT, triggering for time series). On the
supervision side we examine \textbf{labels and uncertainty} ---
inter-annotator variance, probabilistic labels, and a Bayesian view of
priors, likelihoods, and posteriors --- and then formalize the
\textbf{bias--variance} tradeoff with parsimony and regularization. A
major focus is \textbf{Data Leakage} in materials workflows
(pre-processing, temporal, and group/spatial), tackled with proper
holdout, K-fold, LOOCV, and stratified validation. We close with the
\textbf{error measures} that decide what ``good'' actually means:
MAE/MSE/RMSE and \(R^2\) for regression, and confusion matrices,
precision/recall, F1/Dice, IoU, and categorical cross-entropy for
classification and segmentation.

\textbf{Exercise:}\\
Construct a deliberately flawed ML pipeline and diagnose its failure.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsubsection{Unit II --- Representation Learning for Microstructures
(Weeks
4--6)}\label{unit-ii-representation-learning-for-microstructures-weeks-46}

\emph{(Aligned with early neural networks in MFML)}

\paragraph{Week 4 -- From classical microstructure metrics to learned
representations}\label{week-4-from-classical-microstructure-metrics-to-learned-representations}

\emph{Lecture: Tuesday, 05.05.2026, 14:15-15:45 \textbar{} Exercise:
Thursday, 07.05.2026, 16:15-17:45}

\begin{itemize}
\tightlist
\item
  Grain size, phase fractions, orientation maps.
\item
  Limits of hand-crafted microstructure features.
\item
  Transition to learned representations.
\end{itemize}

\textbf{Summary:} This unit marks the transition from classical,
hand-crafted microstructure quantification (like grain size and phase
fractions) to the modern paradigm of \textbf{learned representations}.
We first review traditional stereological metrics and their limitations
in capturing complex structural nuances. We then introduce the
foundational unit of modern ML: the \textbf{artificial neuron}. By
understanding weights, biases, and non-linear activation functions, we
build the framework for Multi-Layer Perceptrons (MLPs) that can
automatically learn optimal features from materials data.

\textbf{Exercise:}\\
Compare classical features vs simple NN-based features for
microstructure tasks.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\paragraph{Week 5 -- Neural networks for microstructure
images}\label{week-5-neural-networks-for-microstructure-images}

\emph{Lecture: Tuesday, 12.05.2026, 14:15-15:45 \textbar{} Exercise:
Thursday, 14.05.2026, 16:15-17:45 (cancelled - Himmelfahrt)}

\begin{itemize}
\tightlist
\item
  CNN intuition: filters as structure detectors.
\item
  Example tasks: phase segmentation, defect detection, porosity
  identification.
\item
  Overfitting risks with small datasets.
\end{itemize}

\textbf{Summary:} This unit introduces \textbf{Convolutional Neural
Networks (CNNs)}, the workhorse of modern computer vision, and applies
them to materials characterization. We explore how convolutions allow
networks to automatically learn hierarchical structure detectors---from
simple edges to complex phase morphologies---while drastically reducing
the number of parameters compared to standard MLPs. Through case studies
in phase segmentation and defect detection, students learn the intuition
behind filters, pooling, and the unique challenges of applying deep
learning to high-resolution, noisy experimental micrographs.

\textbf{Exercise:}\\
Train a small CNN on microstructure images; analyze failure cases.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\paragraph{Week 6 -- Data scarcity \& transfer
learning}\label{week-6-data-scarcity-transfer-learning}

\emph{Lecture: Tuesday, 19.05.2026, 14:15-15:45 \textbar{} Exercise:
Thursday, 21.05.2026, 16:15-17:45}

\begin{itemize}
\tightlist
\item
  Why materials datasets are small.
\item
  Transfer learning from natural images vs self-supervised pretraining.
\item
  When transfer learning helps---and when it does not.
\end{itemize}

\textbf{Summary:} This unit addresses the fundamental bottleneck of
materials informatics: \textbf{Data Scarcity}. We explore how to build
powerful deep learning models when only a few hundred labeled images or
signals are available. The core focus is on \textbf{Transfer Learning},
where we leverage knowledge from models pretrained on millions of
natural images to accelerate learning and improve generalization on
materials tasks. We also cover \textbf{Data Augmentation} strategies
tailored for scientific data and discuss when and why transferring
knowledge across different physical domains succeeds or fails.

\textbf{Exercise:}\\
Fine-tune a pretrained model; compare against training from scratch.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsubsection{Unit III --- Learning from Processing Data (Weeks
7--9)}\label{unit-iii-learning-from-processing-data-weeks-79}

\paragraph{Week 7 -- Time-series and process
monitoring}\label{week-7-time-series-and-process-monitoring}

\emph{Lecture: Tuesday, 26.05.2026, 14:15-15:45 \textbar{} Exercise:
Thursday, 28.05.2026, 16:15-17:45}

\begin{itemize}
\tightlist
\item
  Processing signals: temperature cycles, AM melt pool signals, SPS,
  rolling.
\item
  Regression and sequence models as surrogates.
\item
  Relation to MFML concepts of generalization.
\end{itemize}

\textbf{Summary:} This unit explores the application of machine learning
to \textbf{Time-Series Data}, specifically for monitoring and predicting
materials processing outcomes. We introduce \textbf{Recurrent Neural
Networks (RNNs)} and their advanced variants like \textbf{LSTMs}, which
are designed to handle sequential dependencies. We discuss the critical
preprocessing steps of signal smoothing and triggering required to
handle noisy experimental logs. Through case studies in additive
manufacturing and process stability, students learn how to build models
that ``remember'' the processing history to predict future states and
detect anomalies in real-time.

\textbf{Exercise:}\\
Predict a process outcome from time-series data using regression or
simple RNNs.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\paragraph{Week 8 -- Generalization, robustness, and process
windows}\label{week-8-generalization-robustness-and-process-windows}

\emph{Lecture: Tuesday, 02.06.2026, 14:15-15:45 \textbar{} Exercise:
Thursday, 04.06.2026, 16:15-17:45 (cancelled - Fronleichnam)}

\begin{itemize}
\tightlist
\item
  Sensitivity to noise and parameter drift.
\item
  Overfitting in process--property models.
\item
  Robustness as a design criterion.
\end{itemize}

\textbf{Summary:} This unit shifts the focus from model performance to
\textbf{Model Reliability}. We explore the Bias-Variance tradeoff and
the fundamental challenge of generalization---ensuring that an ML model
works on new, unseen data from the factory floor. We introduce robust
validation techniques like K-Fold and Stratified Cross-Validation to
stabilize performance estimates on small materials datasets. A key focus
is on \textbf{Process Robustness}, where we use sensitivity analysis to
identify ``Process Windows''---regions in parameter space where material
quality is maximized and insensitive to industrial noise.

\textbf{Exercise:}\\
Analyze model robustness under perturbed process conditions.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\paragraph{Week 9 -- Inverse problems and process
maps}\label{week-9-inverse-problems-and-process-maps}

\emph{Lecture: Tuesday, 09.06.2026, 14:15-15:45 \textbar{} Exercise:
Thursday, 11.06.2026, 16:15-17:45}

\begin{itemize}
\tightlist
\item
  Process → structure inverse problems.
\item
  ML-guided process maps (e.g.~AM laser power vs scan speed).
\item
  Physics-informed vs unconstrained regression.
\end{itemize}

\textbf{Summary:} This unit explores \textbf{Inverse Problems}---the
cornerstone of materials design where we seek the processing parameters
required to achieve a target microstructure or performance. We contrast
these with causal forward problems and discuss why they are often
ill-posed and multi-valued. We introduce \textbf{Physics-Informed
Learning} as a way to solve these challenges by enriching models with
physical transformations and constraints. Students learn how to build
and interpret \textbf{Process Maps} and ``Process Corridors,'' using
machine learning to visualize safe operating regions in complex
experimental spaces.

\textbf{Exercise:}\\
Construct a simple ML-based process map; compare constrained vs
unconstrained models.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsubsection{Unit IV --- Uncertainty, Surrogates, and Automation
(Weeks
10--12)}\label{unit-iv-uncertainty-surrogates-and-automation-weeks-1012}

\paragraph{Week 10 -- ML for characterization
signals}\label{week-10-ml-for-characterization-signals}

\emph{Lecture: Tuesday, 16.06.2026, 14:15-15:45 \textbar{} Exercise:
Thursday, 18.06.2026, 16:15-17:45}

\begin{itemize}
\tightlist
\item
  Spectral data: XRD, EELS, EDS.
\item
  Denoising, peak finding, dimensionality reduction.
\item
  Using ML without destroying physical meaning.
\end{itemize}

\textbf{Summary:} This unit focuses on the processing of
high-dimensional \textbf{Characterization Signals} (like XRD, EDS, and
EELS) using unsupervised learning. We introduce \textbf{K-Means
Clustering} and \textbf{t-SNE} for the automatic identification and
visualization of phases in large experimental libraries. We then explore
\textbf{Autoencoders}---neural networks that learn to compress complex
spectra into a low-dimensional ``latent space.'' This allows for
advanced denoising and feature extraction, enabling scientists to handle
the massive data volumes produced by modern high-throughput
characterization tools without losing physical insight.

\textbf{Exercise:}\\
Apply PCA/NMF to spectral datasets; interpret components physically.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\paragraph{Week 11 -- Automation in microscopy and
characterization}\label{week-11-automation-in-microscopy-and-characterization}

\emph{Lecture: Tuesday, 23.06.2026, 14:15-15:45 \textbar{} Exercise:
Thursday, 25.06.2026, 16:15-17:45}

\begin{itemize}
\tightlist
\item
  Autofocus, drift correction, parameter selection.
\item
  ML as a control component, not just a predictor.
\end{itemize}

\textbf{Exercise:}\\
Implement a simple ML-assisted autofocus or defect detector.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\paragraph{Week 12 -- Uncertainty-aware regression \& Gaussian
Processes}\label{week-12-uncertainty-aware-regression-gaussian-processes}

\emph{Lecture: Tuesday, 30.06.2026, 14:15-15:45 \textbar{} Exercise:
Thursday, 02.07.2026, 16:15-17:45}

\begin{itemize}
\tightlist
\item
  Aleatoric vs epistemic uncertainty in experiments.
\item
  Gaussian Processes as uncertainty-aware surrogates.
\item
  Exploration vs exploitation in experimental design.
\item
  Connection to materials acceleration platforms.
\end{itemize}

\textbf{Exercise:}\\
Compare GP regression and NN ensembles for a process-parameter problem.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsubsection{Unit V --- Physics, Trust, and Synthesis (Weeks
13--14)}\label{unit-v-physics-trust-and-synthesis-weeks-1314}

\paragraph{Week 13 -- Physics-informed and constrained
ML}\label{week-13-physics-informed-and-constrained-ml}

\emph{Lecture: Tuesday, 07.07.2026, 14:15-15:45 \textbar{} Exercise:
Thursday, 09.07.2026, 16:15-17:45}

\begin{itemize}
\tightlist
\item
  Embedding physical constraints into ML models.
\item
  Penalty terms, soft constraints, hybrid approaches.
\item
  Failure modes of unconstrained models.
\end{itemize}

\textbf{Exercise:}\\
Train a constrained model for a processing or characterization task.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\paragraph{Week 14 -- Integration, limits, and
reflection}\label{week-14-integration-limits-and-reflection}

\emph{Lecture: Tuesday, 14.07.2026, 14:15-15:45 \textbar{} Exercise:
Thursday, 16.07.2026, 16:15-17:45}

\begin{itemize}
\tightlist
\item
  Explainability for experimental ML (CAMs, SHAP).
\item
  Why ML fails in real labs.
\item
  Where ML genuinely changes materials processing.
\end{itemize}

\textbf{Exercise:}\\
Mini-project presentations and critical discussion.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{Learning Outcomes}\label{learning-outcomes}

Students completing this course will be able to:

\begin{itemize}
\tightlist
\item
  Interpret materials processing and characterization data as learning
  problems.
\item
  Build ML pipelines for microstructure analysis, process prediction,
  and spectral data.
\item
  Understand the physics of data formation to avoid common ML pitfalls.
\item
  Evaluate generalization, robustness, and uncertainty in experimental
  ML models.
\item
  Apply Gaussian Processes and neural networks as surrogate models.
\item
  Integrate physical constraints into ML workflows.
\item
  Critically assess claims about ML in materials processing and
  characterization.
\end{itemize}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{Lab Possibilities}\label{lab-possibilities}

\begin{itemize}
\tightlist
\item
  Microscopy datasets: noise, metadata, units, and artifacts.
\item
  Fourier inspection of SEM/TEM images.
\item
  Broken vs correct ML pipelines (data leakage case studies).
\item
  Feature extraction vs learned representations.
\item
  Fine-tuning pretrained CNNs on microstructures.
\item
  Process--property regression with uncertainty.
\item
  GP-based process maps.
\item
  Spectral decomposition (NMF) of EELS/XRD data.
\item
  ML-assisted autofocus or EBSD pattern classification.
\item
  Multi-modal fusion of images, spectra, and process parameters.
\end{itemize}

\textbf{Summary:} This unit explores the cutting edge of
\textbf{Autonomous Characterization}, where machine learning moves from
passive data analysis to active instrument control. We introduce
\textbf{Multi-Modal Data Fusion} techniques to combine information from
diverse sensors like SEM images, EDS spectra, and process logs using
Bayesian frameworks. We then discuss \textbf{Reinforcement Learning
(RL)} as a tool for automating complex laboratory tasks, such as
instrument tuning and process optimization. Through case studies in
microscopy and industrial processing, students learn how to build
integrated pipelines that can autonomously find, characterize, and
decide the next steps of an experiment.

\protect\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-sandfeld2024materials}
Sandfeld, S. (2024). \emph{Materials data science: Introduction to data
mining, machine learning, and data-driven predictions for materials
science and engineering}. Springer Nature.

\end{CSLReferences}




\end{document}
